---
title: "R Basics Playground"
author: "Oskar Hagen"
execute:
  freeze: auto
editor:
  markdown:
    wrap: 72
---

# Welcome

This is an R Tutorial. We will learn the basics with tree measurements like height and DBH (diameter at breast height).

::: callout-tip
Run code blocks in RStudio or directly in R terminal. Lines that start with `#` are comments.
:::

## Learning goals

-   Create variables and do simple math
-   Work with vectors and data frames
-   Use functions and write your own
-   Make a basic plot
-   Practice with mini exercises and small challenges

# 1. Hello, R

```{r}
# A comment starts with #
print("Hello, R!")
```

# 2. Variables and math (tree height)

Imagine you measured a tree height in meters.

```{r, eval=F}
# Create variables
height_m <- 18.5
extra_m <- 1.2

# Basic math
height_m + extra_m
height_m - extra_m
height_m * 2
height_m / 2
height_m ^ 2
```

::: callout-note
In R, `<-` is the classic assignment arrow. You can also use `=` but `<-` is more common.
:::

## Quick exercise

1.  Make a variable called `dbh_cm` with a DBH value (in centimeters).
2.  Make a variable called `plot_id` with a number.
3.  Add 5 to `dbh_cm` and print the result.

# 3. Vectors (many trees)

Vectors are lists of values of the same type.

```{r, eval=FALSE}
# A numeric vector of DBH measurements (cm)
dbh_cm <- c(12, 18, 22, 15, 30)

# A character vector of tree species
species <- c("oak", "maple", "pine", "birch", "cedar")

# Access by position (1-based)
dbh_cm[1]

# Basic stats
mean(dbh_cm)
max(dbh_cm)

```

## Mini exercise

-   Create a vector called `heights_m` with 5 tree heights.
-   Compute the mean and the max.
-   Add a new height to the end.

# 4. Data frames (a simple field sheet)

Data frames are tables. Each column can be a different type.

```{r, eval=FALSE}
trees <- data.frame(
  tree_id = c("T1", "T2", "T3", "T4"),
  species = c("oak", "maple", "pine", "cedar"),
  dbh_cm = c(25, 18, 30, 14),
  height_m = c(19, 16, 23, 12),
  healthy = c(TRUE, FALSE, F, T)
)

trees

```

:::{.callout-tip icon="🧪"}

### Single Exercise: Selecting columns and rows

Select the `species` column, the first row, and all trees with DBH greater than 20 cm.

```{r, eval=FALSE}
trees$species
trees[1, ]
trees[trees$dbh_cm > 20, ]

```
**Extra** Think of alternative ways of doing this
 
:::

# 5. Functions

Functions let you reuse code.

```{r, eval=FALSE}
# Convert centimeters to inches
cm_to_in <- function(cm) {
  cm / 2.54
}

cm_to_in(30)

```


::: {.callout-tip icon="🧪"}

### Group exercise: Create a function that converts m to cm (10 min)

1. Create a function named `m_to_cm` that takes a numeric input in meters and returns the value in centimeters.
  use 'function()' to define the function and remember that 1 meter = 100 centimeters.
2. Test the function with a few values (e.g., 1.5 m, 2.3 m, 0.75 m).
  call the function with single or vector inputs to see the results.

#### Extras

1. Convert the column `height_m` in the `trees` data frame to centimeters using your function and add it as a new column `height_cm`.
  use '$' to add the new column to the data frame.

:::

# 6. Simple plots (DBH vs height)

```{r, eval=FALSE}
# Basic scatter plot
plot(trees$dbh_cm, trees$height_m,
     main = "Tree Height vs DBH",
     xlab = "DBH (cm)", ylab = "Height (m)")
```

::: callout-tip
Try changing colors with `col = "forestgreen"` or `pch = 19`.
:::

# 7. Putting it together (a small calculation)

A tiny calculation you might do in the field is basal area. Here is a very simple version using DBH in centimeters.

```{r eval=F}
# Basal area in square meters for each tree
# area = pi * (radius_m)^2
radius_m <- (trees$dbh_cm / 100) / 2
basal_area_m2 <- pi * radius_m ^ 2

# Add to the data frame
trees$basal_area_m2 <- basal_area_m2

trees
```

::: {.callout-tip icon="🧪"}

### Group exercise: Do a scatter plot of DBH and Height (10 min)

1.  Make a vector of 4 DBH values and find the median.
  use 'c()' to create the vector and 'median()' to find the median.
2.  Create a data frame called 'mytrees' with two columns: `tree_id` and `height_cm` (4 rows).
  use 'data.frame()' to create the data frame.
3.  Merge the DBH vector from step 1 into the data frame as a new column `dbh_cm`.
  use '$' to add the new column. You can also use 'cbind()' to combine the data frame and vector.
4.  adds 160 cm to all height_cm values.
  use '+' to add 160 to the height_cm column. Remember R likes to vectorize!

#### Extras

1. Write a function that adds 160 cm to a height_cm value.
  define a function using 'function()' that takes a DBH value and returns the value plus 10.
2. **Random sample**: Use `sample()` to pick 2 random trees from `mytrees$tree_id`.

:::

::: callout-note
If you get stuck, run code line by line and use '?' to get help, e.g., `?median`.
:::

::: {.callout-tip icon="🧪"}

### Group exercise: Loop over your entire data frame and print each row (10 min)

```{r, eval=FALSE}
for (i in 1:nrow(mytrees)) {
  print(mytrees[i, ])
}

```
#### Extras

Modify the loop to print only trees with DBH greater than the median DBH.
  Use `if` statement inside the loop to check the condition before printing.

:::


### Mergin all we learned today

```{r, eval=FALSE}
install.packages("here")
library(here)
#list files ending in cvs in working directory
csvs <- list.files(path = here("./data/raw/PSP/"), pattern = "\\.csv$", full.names = TRUE)

# load each csv and merge into a single data frame
all_data <- do.call(rbind, lapply(csvs, read.csv))

# the same as above but more explicit using loops
all_data <- data.frame()
for (file in csvs) {
  #get file name from file
  f_name <- basename(file)
  # remove .csv at the end and PSP_
  f_name <- gsub("^PSP_|\\.csv$", "", f_name)
  print(f_name)
  # read csv
  temp_data <- read.csv(file)
  # add new column with file name
  temp_data$PSP <- f_name
  print(dim(temp_data))
  all_data <- rbind(all_data, temp_data)
}
dim(all_data)

str(all_data)
# spot values that can not be converted to numeric in DBH or Height
mask <- is.na(as.numeric(all_data$DBH))
all_data[mask,]

# manual fix of DBH
# there was text instead of number for Tag.Number 1234 in PSP NPD42_OH on DBH. i.e. 1o.4
all_data[all_data$Tag.Number==1234&all_data$PSP=="NPD42_OH", "DBH"] <- 10.4


# convert DBH to numeric
all_data$Height <- as.numeric(all_data$Height)

# plot scatterplot dbh vs height for each PSP
plot(all_data$DBH, all_data$Height,
     main = "Tree Height vs DBH",
     xlab = "DBH (cm)", ylab = "Height (m)", col = as.factor(all_data$PSP))
# add legend for PSP
legend("topleft", legend = unique(all_data$PSP), col = 1:length(unique(all_data$PSP)), pch = 1)

```

::: {.callout-tip icon="🧪"}

### Extra Solo exercise: check outliers

1. Identify trees with height smaller than 200 cm.
  Use logical indexing to filter rows where height > 100. (e.g. all_data[all_data$Height<200,])

2. Remove the outlines from the data frame, you talked with people that collected the data first! ;)
  Use logical indexing to keep only rows where height >= 100, or remove rows with height < 100.
  you can either use the function 'which', or use logical indexing directly. Remember '!' means NOT in R, thus a vector x <- c(F, T, F), then !x is c(T, F, T)
  
3. Remove any tree that is deadwood from the data frame.
  Use logical indexing to keep only rows where Deadwood column is FALSE.
  
```{r, eval=FALSE}
# remove any entry with Species containing "Dead wood"
data_no_DeadWood <- all_data[!grepl("Dead wood", all_data$Species), ]
# alternatively, remove any entry on Species containing "Dead wood (CLASS 1)" or "Dead wood (CLASS 2)" using %in%
data_no_DeadWood <- all_data[!(all_data$Species %in% c("Dead wood (CLASS 1)", "Dead wood (CLASS 2)")), ]

```

Hint: remove the '!' to check the selection before you remove. Also use 'dim()' to check the size of the data frame before and after removing rows.

:::


# 8. Keeping it clean

In order to keep your workflow clean, it is recomended that intermediate results are saved to files. This way you can always go back to a previous step without re-running everything. 

Also, save the scripts in separate steps, e.g. `01_load_data.R`, `02_clean_data.R`, `03_analyze_data.R`, etc.

```{r, eval=FALSE}
# save cleaned data to csv
write.csv(data_no_DeadWood, file = here("./data/cleaned_tree_data.csv"), row.names = FALSE)
```